Seminar 4: Two-Sample Testing
========================================================
We will begin by loading necessary packages and the photorec dataset

```{r Loading packages and datasets}
library(plyr)
library(lattice)
prDat <- read.table("GSE4051_data.tsv")
str(prDat, max.level=0)
prDes <- readRDS("GSE4051_design.rds")
str(prDes)
```

Let's extract the gene expression data for 100 random probes and place the data into a separate dataframe

```{r Extracting 100 random probes and placing data into new dataframe}
set.seed(16)
random.probes <- sample(1:nrow(prDat),100)
mini.df <- prDat[random.probes,]
mini.df <- data.frame(gExp = as.vector(t(as.matrix(mini.df))), gene = factor(rep(rownames(mini.df), each = ncol(mini.df)), levels = rownames(mini.df)))
mini.df <- suppressWarnings(data.frame(prDes, mini.df))
str(mini.df)
```
Create a function that computes a t-test, Wilcoxon and KS test and returns their respective p-values as a matrix

```{r Two-sample testing by 3 methods function}
ppval <- suppressWarnings(ddply(mini.df, ~ gene, function(x) {
  tt <- t.test(gExp ~ gType, x)
  wt <- wilcox.test(gExp ~ gType, x)
  ks <- with(x, ks.test(gExp[gType=="NrlKO"], gExp[gType=="wt"]))
  round(c(t.test = tt$p.value, Wilcoxon = wt$p.value, KS = ks$p.value), 4)
}))
head(ppval)
```
Plot scatterplots of  p-values (untransformed and log-transformed) from the 3 different statistical tests

```{r fig.width=7, fig.height=6}
plot(ppval[,-1], col="blue")
plot(log(ppval[,-1]), col="red")
```
Set threshold of significance at alpha = 0.05 and count how many probes are "statistically significant hits" by all 3 methods, 2 methods, 1 method or none
```{r Counting number of probes that are "hits"}
alpha  <- ppval[,-1] <= 0.05
apply(alpha,2,sum)
(count.hits <- as.matrix(count(apply(alpha,1,sum))))
```

