Seminar 8: Methylation
========================================================

In this seminar, we are going to perform differential methylation analysis (run on the Illumina Infinium HumanMethylation 450K array) for Acute lymphoblastic leukemia (ALL) cases and healthy B cells as control group. The datasets we are going to use are: GSE39141 which contains 29 ALL bone marrow samples and 4 healthy B cell samples, and to make up for the small sample size for healthy cells, we will also use GSE42865 which contains 9 healthy B cell samples and 7 samples with other conditions. 

## Load 450K methylation array data
First, let's retrieve our datasets from GEO with getGEO from GEOquery package. 

```{r, load.packages, include=FALSE}
library(GEOquery)
library(wateRmelon)
library(IlluminaHumanMethylation450k.db)
```

```{r, load.data}
if (file.exists("methyl_ALL.Rdata")) {
  # if previously downloaded
  load("methyl_ALL.Rdata")
} else {
  # if downloading for the first time
  GSE39141 <- getGEO("GSE39141")
  show(GSE39141)  ## 33 samples (29 ALL and 4 healthy B cells)
  GSE42865 <- getGEO("GSE42865")  # took ~2 mins for JB
  show(GSE42865)  ## 16 samples (9 healthy cells B cells and 7 other cells)
  
  # Extract expression matrices (turn into data frames at once)
  ALL.dat <- as.data.frame(exprs(GSE39141[[1]]))
  CTRL.dat <- as.data.frame(exprs(GSE42865[[1]]))
  
  # Obtain the meta-data for the samples and rename them perhaps?
  ALL.meta <- pData(phenoData(GSE39141[[1]]))
  CTRL.meta <- pData(phenoData(GSE42865[[1]]))
  
  # create some labels
  ALL.meta$Group <- c(rep("ALL", 29), rep("HBC", 4))
  ## ALL: Case; HBC: Healthy B Cells
  
  # Subset both meta-data and data for control (healthy) donors
  CTRL.meta <- droplevels(subset(CTRL.meta, grepl("Healthy donor", characteristics_ch1.1)))
  CTRL.dat <- subset(CTRL.dat, select = as.character(CTRL.meta$geo_accession))
  
  # Rename variables
  names(ALL.dat) <- paste(ALL.meta$Group, gsub("GSM", "", names(ALL.dat)), sep = "_")
  names(CTRL.dat) <- paste("HBC", gsub("GSM", "", names(CTRL.dat)), sep = "_")
  
  # save the data to avoid future re-downloading
  save(ALL.dat, CTRL.dat, ALL.meta, CTRL.meta, file = "methyl_ALL.Rdata")
}
```
Let's plot the average beta distribution of the ALL samples and the CTRL samples separately. 

```{r}
library(ggplot2)
avg.betas <- c(rowMeans(ALL.dat, na.rm = T), rowMeans(CTRL.dat, na.rm = T))
plotDat <- data.frame(Beta = avg.betas,
                      Dataset = rep(c('ALL', 'CTRL'), each = nrow(ALL.dat)))
(probeAvg <- ggplot(data = plotDat, aes(x = Beta, col = Dataset)) +
   geom_density() +
   ggtitle("Average Beta distribution of ALL vs CTRL samples") +
   xlab("Beta values") +
   ylab("Density") +
   theme_bw()
)
```
## Normalization

Since the distribution of average beta values between the two experiments are not the same, we will need to normalize the data. 
```{r normalize.data}
# combine data from two experiments into one matrix, each column represents
# beta values of one sample
beta.matrix <- as.matrix(cbind(ALL.dat, CTRL.dat))
str(beta.matrix, max.level = 0)
system.time(beta.norm <- betaqn(beta.matrix))
```
Let's compare the beta value distributions of the two experiments before and after normalization. 
```{r}
dat.probeMeans <- c(rowMeans(beta.norm[, 1:ncol(ALL.dat)], na.rm = TRUE),
                    rowMeans(beta.norm[, ncol(ALL.dat):ncol(beta.norm)], na.rm = TRUE))
plotNorm <-
  rbind(data.frame(plotDat, Norm = "Before"),
        data.frame(Beta = dat.probeMeans,
                   Dataset = rep(c('ALL', 'CTRL'), each = nrow(ALL.dat)),
                   Norm = "After"))
plotNorm$Norm <- factor(plotNorm$Norm, levels = c("Before", "After"))
(probeAvgNorm <- ggplot(data = plotNorm, aes(x = Beta, col = Dataset)) +
   geom_density() +
   facet_grid(Norm ~ .) +
   ggtitle("Density of Beta values before and after normalization") +
   xlab("Beta") +
   ylab("Density") +
   theme_bw()
)
```
## M-values
We can see above that the beta value distribution is predominantly bimodal on the interval [0,1]. This type of distribution is not very compatible with the typical assumptions of linear models. To circumvent this problem, we will compute M-values by applying a logit transformation on the data to convert it to a continuous variable that spans (−∞,∞). 
```{r M.values}
M.norm <- beta2m(beta.norm)
```
## CpG Islands
We will aggregate our data to focus on probes that reside in CpG islands (CGIs) and then detect differentially methylated CGIs. 
```{r}
# Extracting probe ID to CpG islands association
cginame <- as.data.frame(IlluminaHumanMethylation450kCPGINAME)
names(cginame) <- c("Probe_ID", "cginame")
rownames(cginame) <- cginame$Probe_ID
length(levels(factor(cginame$cginame)))  # No. of CGIs

# restrict probes to those within CGIs
beta.inCGI <- beta.norm[cginame$Probe_ID, ]
M.inCGI <- M.norm[cginame$Probe_ID, ]
nrow(M.inCGI)  # No. of probes within CGIs

# aggregate probes to CGIs
beta.CGI <- aggregate(beta.inCGI, by = list(cginame$cginame), mean, na.rm = T)
rownames(beta.CGI) <- beta.CGI[, "Group.1"]
beta.CGI <- subset(beta.CGI, select = -Group.1)
str(beta.CGI, max.level = 0)
M.CGI <- aggregate(M.inCGI, by = list(cginame$cginame), mean, na.rm = T)
rownames(M.CGI) <- M.CGI[, "Group.1"]
M.CGI <- subset(M.CGI, select = -Group.1)
str(M.CGI, max.level = 0)
```
Checking the relative distribution of CGI M-values:
```{r, warning=FALSE}
# check the distribution of CGI M values with boxplot
library(reshape2)
M.CGI.tall <- melt(t(M.CGI), value.name = 'M', varnames = c('Sample', 'CGI'))
M.CGI.tall$Group <- substring(M.CGI.tall$Sample, 1, 3)
(M.boxplot <- ggplot(data = M.CGI.tall, aes(Sample, M, colour=Group)) +
   geom_boxplot() +
   ggtitle("Distribution of CGI M values") +
   xlab("Samples") +
   ylab("M values") +
   theme_bw() +
   scale_x_discrete(labels = NULL))
```
## Differential methylation analysis with limma
We willuse a linear model to identify differentially methylated CGIs with limma. 

```{r}
library(limma)
design <- data.frame(Group = relevel(factor(gsub("_[0-9]+", "", colnames(M.CGI))), 
    ref = "HBC"), row.names = colnames(M.CGI))
str(design)
(DesMat <- model.matrix(~Group, design))
DMRfit <- lmFit(M.CGI, DesMat)
DMRfitEb <- eBayes(DMRfit)
cutoff <- 0.01
DMR <- topTable(DMRfitEb, coef = "GroupALL", number = Inf, p.value = cutoff)
head(DMR)  # top hits 
```
Using a FDR cutoff of 0.01, we identified 4115 CGIs that are differentially methylated between ALL and control samples. 

Create a heatmap of beta values of top 100 hits:
```{r}
library(gplots)
DMR100 <- topTable(DMRfitEb, coef = "GroupALL", number = 100)
DMR.CGI <- t(as.matrix(subset(beta.CGI, rownames(beta.CGI) %in% rownames(DMR100))))
str(DMR.CGI, max.level = 0)
col <- c(rep("darkgoldenrod1", times = nrow(DMR.CGI)))
col[grepl("HBC", rownames(DMR.CGI))] <- "forestgreen"
op <- par(mai = rep(0.5, 4))
heatmap.2(DMR.CGI, col = redblue(256), RowSideColors = col, density.info = "none", 
    trace = "none", Rowv = TRUE, Colv = TRUE, labCol = FALSE, labRow = FALSE, 
    dendrogram = "row", margins = c(1, 5))
legend("topright", c("ALL", "HBC"), col = c("darkgoldenrod1", "forestgreen"), 
    pch = 15)
```
Create a stripplot of the beta values of the probes within the top CGI hits:
```{r}
DMR5 <- topTable(DMRfitEb, coef = "GroupALL", number = 5)
beta.DMR5probe <- beta.inCGI[cginame[rownames(beta.inCGI), ]$cginame %in% rownames(DMR5), 
    ]
beta.DMR5probe.tall <- melt(beta.DMR5probe, value.name = "M", varnames = c("Probe_ID", 
    "Sample"))
beta.DMR5probe.tall$Group <- factor(gsub("_[0-9]+", "", beta.DMR5probe.tall$Sample))
beta.DMR5probe.tall$CGI <- factor(cginame[as.character(beta.DMR5probe.tall$Probe_ID), 
    ]$cginame)
(beta.DMR5.stripplot <- ggplot(data = beta.DMR5probe.tall, aes(x = Group, y = M, 
    color = Group)) + geom_point(position = position_jitter(width = 0.05), na.rm = T) + 
    stat_summary(fun.y = mean, aes(group = 1), geom = "line", color = "black") + 
    facet_grid(. ~ CGI) + ggtitle("Probe beta values within top 5 DM CGIs") + 
    xlab("Group") + ylab("beta") + theme_bw())
```
Plot location of differential methylated probes along each chromosome:
```{r}
# get the length of chromosome 1-22 and X
chrlen <- unlist(as.list(IlluminaHumanMethylation450kCHRLENGTHS)[c(as.character(1:22), 
    "X")])
chrlen <- data.frame(chr = factor(names(chrlen)), length = chrlen)
chr <- IlluminaHumanMethylation450kCHR  # get the chromosome of each probe
# get the probe identifiers that are mapped to chromosome
chr <- unlist(as.list(chr[mappedkeys(chr)]))
# get chromosome coordinate of each probe
coord <- IlluminaHumanMethylation450kCPGCOORDINATE
# get the probe identifiers that are mapped to coordinate
coord <- unlist(as.list(coord[mappedkeys(coord)]))
coord <- data.frame(chr = chr[intersect(names(chr), names(coord))], coord = coord[intersect(names(chr), 
    names(coord))])
# coordinates of probes in DM CGIs
coordDMRprobe <- droplevels(na.omit(coord[cginame[cginame$cginame %in% rownames(DMR), 
    ]$Probe_ID, ]))
(coord.plot <- ggplot(data = coordDMRprobe) + geom_linerange(aes(factor(chr, 
    levels = c("X", as.character(22:1))), ymin = 0, ymax = length), data = chrlen, 
    alpha = 0.5) + geom_point(aes(x = factor(chr, levels = c("X", as.character(22:1))), 
    y = coord), position = position_jitter(width = 0.03), na.rm = T) + ggtitle("DMR positions on chromosomes") + 
    ylab("Position of DMRs") + xlab("chr") + coord_flip() + theme_bw())
```



