Seminar 2: Coin Toss Simulator
========================================================

We will simulate flipping a coin n times, where the coin comes up heads with a probability p (say, 0.3) and plot the proportion of heads. 

```{r coinFlips}
p  <- 0.3 
n <- seq(from=1, to=750, by=1)
coinFlips <- sapply(n, rbinom, size=1, prob=p)
count_heads <- sapply(coinFlips, sum) 
compute_proportions <- count_heads/n 
```

```{r fig.width=7, fig.height=6}
plot(n,compute_proportions, type="l", xlab= "n (sample size)", ylab="Proportion of heads", main="Coin Toss Simulator") # plot the proportion of heads against n
abline(h=p, col="purple", lwd=2) 
```

As n increases, proportion of heads tends to p (note that the *true* probability is represented by the purple line).

We can simulate the coin flipping over many trials of varying sample sizes and compute the mean proportion of heads.

```{r Coin Flip Simulation for varying n}
trials <- 500
varying_n <- 10^(1:4)
names(varying_n) <- paste0("n =", varying_n)
coin_tosses <- function(varying_n, trials, prob) {
  mat <- matrix(rbinom(n = trials * varying_n, size = 1, prob = p), nrow = trials, 
              ncol = varying_n)
  prop_heads <- rowMeans(mat)
  return(prop_heads)
}
all_prop <- sapply(varying_n, coin_tosses, trials = trials, p = p)
```

```{r fig.width=7, fig.height=6}
boxplot(all_prop, xlab = "n (Sample Size)", ylab="Proportion of Heads", main="Boxplot of Mean Proportion of Heads over Varying Sample Sizes")
```

As before, we observe that as n increases, the mean proportion of heads tends to p, (consistent with the Weak Law of Large Numbers). 