Seminar 4: Run Time Calculation
===================================================

We would like to calculate the run time needed to perform two-sample testing (for differential expression analysis) using various datasets containing varying numbers of probes (n).

We will begin by loading necessary packages and the photorec dataset

```{r loading necessary packages and dataset}
library(plyr)
library(lattice)
prDat <- read.table("GSE4051_data.tsv")
str(prDat, max.level=0)
prDes <- readRDS("GSE4051_design.rds")
str(prDes)
```

Let's draw increasing numbers (n) of probes from the dataset

```{r Picking increasing n numbers of probes}
n <-  seq(from = 50, to = 1000, by = 100)
set.seed(72)
```
Now we will write a function that prepares a dataframe of n probes and then calculates run time for differential gene expression analysis
```{r Function for calculating run time on various n-sized datasets}
Calc.Run.Time <- function(x){
  random.probes <- sample(1:nrow(prDat),x)
  mini.df <- prDat[random.probes,]
  mini.df <- data.frame(gExp = as.vector(t(as.matrix(mini.df))), gene = factor(rep(rownames(mini.df), each = ncol(mini.df)), levels = rownames(mini.df)))
  mini.df <- suppressWarnings(data.frame(prDes, mini.df))
  run.time <- system.time(ppval <- suppressWarnings(ddply(mini.df, ~ gene, function(w) {
    tt <- t.test(gExp ~ gType, w)
    wt <- wilcox.test(gExp ~ gType, w)
    ks <- with(w, ks.test(gExp[gType=="NrlKO"], gExp[gType=="wt"]))
    round(c(t.test = tt$p.value, Wilcoxon = wt$p.value, KS = ks$p.value), 4)
  })))
  return(data.frame(probes=x, elapsed=run.time["elapsed"]))
}
```
Now let's compute run times for each value of n
```{r Running the function over all values of n and assembling a nice neat dataframe of outputs}
nRunTimes <- lapply(n,Calc.Run.Time)
run.time.df <- do.call(rbind,nRunTimes)
head(run.time.df)
```
Plot the elapsed time against number of probes in dataset

```{r fig.width=7, fig.height=6}
xyplot(elapsed~probes,run.time.df, xlab="Number of probes", ylab="Run time")
```
As we would expect, the run time increases linearly with increasing number of probes in the dataset
